Towards Universal Forest Mapping: Foundation Embeddings Outperform Hand-Crafted Harmonics for Deciduous–Evergreen Classification

### 2025-10-31 article/article.tex
safe save

\subsection{Study area and eco-regional stratification}
\label{sec:greco}
We focus on mainland France and Corsica, where temperate, montane, and Mediterranean forest types co-exist within short distances. To preserve ecological diversity in both training and evaluation we rely on the eleven \emph{Grandes Régions Écologiques} (GRECO) defined by the French National Forest Inventory \citep{IGN2013GRECO}. These eco-regions capture the dominant gradients from Atlantic mixed forests through semi-continental plateaus to Mediterranean evergreen stands and Alpine massifs. We partitioned the national forest mask into 639 non-overlapping 2.5\,km\,$\times$\,2.5\,km tiles (Figure~\ref{fig:training_tiles}), retaining tiles that contain at least ten inventory or plot-labelled pixels and are separated by 5\,km to limit spatial autocorrelation. After the reference quota is met, remaining forest pixels in each tile are populated with buffered BD Forêt V2 polygons so that both feature families share an identical footprint. Sample weights compensate for the unequal forest area of each eco-region: for region \(r\), the weight \(w_r\) equals the ratio between its effective forest area fraction \(A_r/\sum_{j} A_j\) and the observed sample fraction \(n_r/N\). We normalise the weights so that \(\sum_i w_i = N\); the resulting weights range from 0.89 in densely sampled regions to 1.48 in rare mountainous tiles. A small fraction of coastal pixels remains outside the GRECO mask and is excluded from stratified summaries.

@@

\subsection{Evaluation metrics and derived summaries}
Fold-level metrics include overall accuracy, class-specific precision and recall, macro- and weighted-F1, and the four entries of the confusion matrix. We report the mean, standard deviation, and interquartile range across folds, while eco-region summaries average validation folds using the eco-region weights. Calibration is assessed with 10 equal-width probability bins on the evergreen posterior: we quote the expected calibration error (ECE), the maximum calibration error (MCE), and the macro-F1 variation when the decision threshold shifts to 0.45 or 0.55.

Spatial coherence is evaluated on the final mosaics tile by tile. Edge density measures how many metres of deciduous–evergreen boundary lie within a tile, normalised by tile area (km of edge per km$^2$). Connected patch density treats each 8-connected clump of deciduous or evergreen pixels as a patch and counts how many such components occur per 100\,km$^2$ of forest. Fewer patches therefore indicate smoother, more coherent predictions, whereas high densities reveal speckle. We also gauge the effect of simple post-processing by applying a \(3 \times 3\) median filter to the harmonic predictions. Tile-level cross-validation predictions yield accuracy and macro-F1 per tile; we convert the embedding-minus-harmonic deltas to $z$-scores relative to the national distribution, categorising tiles into embedding advantage (\(z \ge +1\sigma\)), rough parity (\(|z| < 1\sigma\)), or harmonic advantage (\(z \le -1\sigma\)). Pearson and Spearman correlations between these deltas and contextual covariates underpin the heterogeneity analysis.

@@

\subsection{Temporal stability evaluation}
Temporal transfer is evaluated by applying the 2023 Emb-14 random forest to the 2022, 2020, and 2018 embedding datasets. The evaluation script reuses the cross-validation manifest so that each pixel inherits its original fold assignment, enabling direct comparison between cross-validation predictions and cross-year inference. For every evaluated year we emit the same artefacts as during training: overall metrics, fold tables, eco-region breakdowns, per-pixel probabilities, and per-tile metrics. Metrics are summarised relative to the 2023 inference baseline and to the 2023 cross-validation reference to quantify year-to-year drift (Supplementary Tables~S4–S5).

\paragraph{Hypotheses.} The experiments test three hypotheses. \textbf{H1} posits that pre-trained embeddings (EMB-14) match or exceed the accuracy and calibration of physics-informed harmonics (HARM-14) under identical data, folds, and classifiers. \textbf{H2} states that embeddings, by encoding spatial context, deliver smoother classification maps and exhibit systematic links with ancillary drivers such as soils, climate, and forest composition. \textbf{H3} asserts that a random forest trained on 2023 embeddings remains robust when applied to other annual embedding vintages and preserves compatibility with legacy national products.

\subsection{Embedding–harmonic similarity analysis}
To probe the relationship between embeddings and explicit phenology we regress each embedding dimension onto the full set of harmonic descriptors using ridge regression,
\begin{equation*}
    \hat{\boldsymbol{\beta}} = \arg\min_{\boldsymbol{\beta}} \big\| \mathbf{y} - \mathbf{X}\boldsymbol{\beta} \big\|_2^2 + \lambda \|\boldsymbol{\beta}\|_2^2,
\end{equation*}
where \(\mathbf{y}\) is the embedding response, \(\mathbf{X}\) the harmonic design matrix, and \(\lambda\) the cross-validated regularisation parameter. Within every eco-region we standardise predictors and responses, adopt a GroupKFold split that keeps tiles intact, and select \(\lambda\) through RidgeCV with up to five folds. Weighted $R^2$ scores are computed per tile and summarised by eco-region; we also catalogue the normalised absolute ridge coefficients and the harmonic feature with highest Pearson correlation for each embedding component. Detailed coefficients, $R^2$ distributions, and tile-level diagnostics appear in Supplementary Section~S3.

@@

\subsection{H2 -- Spatial coherence and landscape context}
Hypothesis H2 examines whether the embedding representation yields smoother outputs and whether these gains relate to ancillary environmental gradients. National edge density drops from 6.69 to 2.76\,km\,km\(^{-2}\) when switching from HARM-14 to EMB-14, a 59\,\% reduction. Patch density declines from 6.81\,k to 2.16\,k components per 100\,km² (−68\,%). The largest smoothing occurs in maritime pine mosaics (Greater Crystalline and Oceanic West), where the harmonic map fragments mixed stands into speckled artefacts. EMB-14 also maintains crisp elevational boundaries in the Alps, Jura, and Vosges, while matching National Forest Inventory proportions (64.4\,\% deciduous, 35.6\,\% evergreen; Figure~\ref{fig:national_map}). Table~\ref{tab:coherence_summary} quantifies the national reduction in edge and patch densities.

Figure~\ref{fig:h2_multiscale} visualises these differences across five representative tiles. Each row pairs a Sentinel-2 chip (left) with the embedding and harmonic predictions (middle and right). The rows cover GRECO letters G (Greater Semi-Continental East), C (Corsica), A (Alps), O (Oceanic Southwest), and the Central Massif. Embeddings preserve large deciduous blocks (amber, \#e3712c) and evergreen belts (azure, \#2693c1) with minimal speckle, while harmonics tend to fragment mixed stands into checkerboards despite sharing the same Random Forest.

@@

Table~\ref{tab:similarity_summary} summarises the regional mean $R^2$ values and dominant cues. The absence of linear reconstruction combined with these alignment patterns supports a hybrid interpretation: AlphaEarth embeddings retain interpretable phenological axes—seasonal vigor and baseline structure—while injecting additional spatial context learned during pre-training. This extra context likely underpins the smoother boundaries and higher accuracy gains observed in mixed Atlantic and montane landscapes, where purely temporal descriptors struggle.

@@

Calibration quality underpins operational usability. EMB-14 reaches an expected calibration error of 0.033 versus 0.059 for HARM-14, and maintains macro-F1 stability when the decision threshold moves within the \([0.45,0.55]\) range (Supplementary Figure~S6). Combined with the reduced speckle visible in Figure~\ref{fig:comparison_products} and the ancillary insights above, these traits simplify both threshold tuning and regional deployment: humid mixed stands can adopt standard thresholds, while dry evergreen regions can either fall back to harmonic cues or tighten evergreen-specific priors without compromising reliability.

>>>>

\subsection{Study area and eco-regional stratification}
\label{sec:greco}
We focus on mainland France and Corsica, where temperate, montane, and Mediterranean forest types co-exist within short distances. To preserve ecological diversity in both training and evaluation we rely on the eleven \emph{Grandes Régions Écologiques} (GRECO) defined by the French National Forest Inventory \citep{IGN2013GRECO}. These eco-regions capture the dominant gradients from Atlantic mixed forests through semi-continental plateaus to Mediterranean evergreen stands and Alpine massifs. We partitioned the national forest mask into 639 non-overlapping 2.5\,km\,$\times$\,2.5\,km tiles (Figure~\ref{fig:training_tiles}), retaining tiles that contain at least ten inventory or plot-labelled pixels and are separated by 5\,km to limit spatial autocorrelation. After the reference quota is met, remaining forest pixels in each tile are populated with buffered BD Forêt V2 polygons so that both feature families share an identical footprint. Sample weights then align the dataset with the actual forest extent of each eco-region. We estimate that extent through an effective forest area \(A_r\), obtained by multiplying each GRECO polygon by the region's mean forest cover ratio provided by the National Forest Inventory, and we sum these areas across regions to get \(\sum_{j} A_j\). The training dataset contains \(N\) labelled forest pixels, with \(n_r\) of them belonging to region \(r\); the per-pixel weight is therefore defined as \(w_r = (A_r / \sum_{j} A_j) / (n_r / N)\). Regions that occupy a large share of forest but are sparse in the sample receive \(w_r > 1\), whereas regions that are already over-represented are down-weighted. For example, the Oceanic Southwest accounts for roughly 22\% of the effective forest area yet only about 21\% of the labelled pixels, so its observations are scaled by \(w_r \approx 1.06\). We normalise the weights so that \(\sum_i w_i = N\); the resulting weights range from 0.89 in densely sampled regions to 1.48 in rare mountainous tiles. A small fraction of coastal pixels remains outside the GRECO mask and is excluded from stratified summaries.

@@

\subsection{Evaluation metrics and derived summaries}
\label{sec:metrics}
Fold-level metrics include overall accuracy, class-specific precision and recall, macro- and weighted-F1, and the four entries of the confusion matrix. Macro-F1 averages the class-specific F1-scores with equal weight so that deciduous and evergreen contribute symmetrically; weighted-F1 applies the empirical class proportions \(n_c/N\) observed in the fold, i.e.
\begin{equation*}
    \mathrm{F1}_{\text{macro}} = \frac{1}{2} \sum_{c \in \{\text{dec}, \text{ever}\}} \mathrm{F1}_c,
    \qquad
    \mathrm{F1}_{\text{weighted}} = \sum_{c \in \{\text{dec}, \text{ever}\}} \frac{n_c}{N} \mathrm{F1}_c,
\end{equation*}
where \(n_c\) is the number of validation pixels of class \(c\) and \(N = n_{\text{dec}} + n_{\text{ever}}\). We report the mean, standard deviation, and interquartile range across folds, while eco-region summaries average validation folds using the eco-region weights. Calibration is assessed with 10 equal-width probability bins on the evergreen posterior, i.e. the random forest probability that a pixel belongs to the evergreen class. For each bin \(b\) with sample set \(S_b\) we compute the mean confidence \(\hat{p}_b = |S_b|^{-1} \sum_{i \in S_b} p_i\) and the observed evergreen fraction \(\hat{y}_b = |S_b|^{-1} \sum_{i \in S_b} \mathbf{1}\{y_i = \text{evergreen}\}\). The expected calibration error is the weighted average of the absolute gaps between these two quantities,
\begin{equation*}
    \mathrm{ECE} = \sum_{b=1}^{10} \frac{|S_b|}{N} \left| \hat{p}_b - \hat{y}_b \right|,
\end{equation*}
with \(N\) the total number of validation pixels, while the maximum calibration error (MCE) records \(\max_b \left| \hat{p}_b - \hat{y}_b \right|\). When a model is perfectly calibrated, \(\hat{p}_b\) and \(\hat{y}_b\) coincide in every bin and both metrics vanish. This reliability analysis clarifies how users can adjust the decision threshold: we also track macro-F1 when the evergreen threshold moves to 0.45 or 0.55 to confirm that modest shifts—such as raising the threshold to suppress evergreen false positives—do not destabilise overall performance.

Spatial coherence is evaluated on the final mosaics tile by tile. Edge density measures how many metres of deciduous–evergreen boundary lie within a tile, normalised by tile area (km of edge per km$^2$). Connected patch density uses an 8-neighbour definition (horizontal, vertical, and diagonal adjacency) to identify every contiguous cluster of forest pixels that share the same class. We label the binary masks for deciduous and evergreen pixels with a \(3\times3\) structure matrix of ones, yielding component counts \(C_{\text{dec}}\) and \(C_{\text{ever}}\). The corresponding densities are
\begin{equation*}
    D_{\text{dec}} = \frac{100\,C_{\text{dec}}}{A}, \qquad
    D_{\text{ever}} = \frac{100\,C_{\text{ever}}}{A}, \qquad
    D_{\text{tot}} = D_{\text{dec}} + D_{\text{ever}},
\end{equation*}
where \(A\) is the forest area (km$^2$) within the tile. When predictions are smooth, large coherent stands dominate and the component counts remain small, producing low densities; salt-and-pepper artefacts fragment stands into many tiny clusters, inflating \(D_{\text{tot}}\). We also gauge the effect of simple post-processing by applying a \(3 \times 3\) median filter to the harmonic predictions. Tile-level cross-validation predictions yield accuracy and macro-F1 per tile; we convert the embedding-minus-harmonic deltas to $z$-scores relative to the national distribution, categorising tiles into embedding advantage (\(z \ge +1\sigma\)), rough parity (\(|z| < 1\sigma\)), or harmonic advantage (\(z \le -1\sigma\)). Pearson and Spearman correlations between these deltas and contextual covariates underpin the heterogeneity analysis.

@@

\subsection{Temporal stability evaluation}
Temporal transfer is evaluated by freezing the 2023 EMB-14 random forest and applying it to the AlphaEarth embedding vintages released for 2022, 2020, and 2018. We reload the cross-validation manifest so that each labelled pixel keeps the fold identifier it had during training. This alignment serves two purposes: (i) it guarantees that the cross-year evaluations use exactly the same supervision set and eco-region weights as the cross-validation runs, and (ii) it allows per-fold comparisons between the original out-of-fold predictions and the cross-year inference.

For every evaluated year we emit the full metric bundle used in training: confusion matrices aggregated over all folds, fold-level tables, eco-region summaries, per-pixel probabilities, and per-tile metrics. Intersection-over-Union (IoU) is derived from the confusion counts for each class \(c \in \{\text{dec}, \text{ever}\}\) as
\begin{equation*}
    \mathrm{IoU}_c = \frac{\mathrm{TP}_c}{\mathrm{TP}_c + \mathrm{FP}_c + \mathrm{FN}_c},
\end{equation*}
with \(\mathrm{TP}_c\), \(\mathrm{FP}_c\), and \(\mathrm{FN}_c\) defined relative to the reference labels (evergreen is the positive class). Macro-IoU averages the two classes with equal weight. IoU quantifies the overlap between the predicted evergreen/deciduous sets and their reference counterparts, complementing the precision/recall perspective. Alongside IoU we report the same accuracy and F1 statistics as for cross-validation; deltas are computed both against the 2023 inference baseline and the 2023 cross-validation reference to characterise year-to-year drift (Supplementary Tables~S4–S5). All artefacts are stored under \texttt{results/evaluation/embeddings\_<year>/}.

\subsection{Embedding–harmonic similarity analysis}
To probe the relationship between embeddings and explicit phenology we regress each embedding dimension onto the full set of harmonic descriptors using ridge regression. For a given eco-region and embedding channel \(k\), we assemble a response vector \(\mathbf{y}^{(k)} \in \mathbb{R}^n\) with one entry per pixel (standardised to zero mean and unit variance) and a design matrix \(\mathbf{X} \in \mathbb{R}^{n \times p}\) containing the \(p\) harmonic descriptors (also z-scored). Ridge regression estimates coefficients \(\hat{\boldsymbol{\beta}}^{(k)}\) by solving
\begin{equation*}
    \hat{\boldsymbol{\beta}}^{(k)} = \arg\min_{\boldsymbol{\beta}} \left\| \mathbf{y}^{(k)} - \mathbf{X}\boldsymbol{\beta} \right\|_2^2 + \lambda \|\boldsymbol{\beta}\|_2^2,
\end{equation*}
where the regularisation strength \(\lambda\) controls the amount of shrinkage applied to the coefficients. We select \(\lambda\) through RidgeCV on a logarithmic grid (20 values between \(10^{-4}\) and \(10^{4}\)), using a GroupKFold split that keeps tiles intact so that no tile contributes to both training and validation folds. Because \(\lambda\) is merely a stabilising hyperparameter—and different embedding channels prefer different values—we do not report it explicitly in the Results; instead we focus on what the fitted models reveal (variance explained and dominant harmonic cues).

After fitting, we compute tile-level coefficients of determination
\begin{equation*}
    R^2_{t,k} = 1 - \frac{\|\mathbf{y}^{(k)}_t - \mathbf{X}_t \hat{\boldsymbol{\beta}}^{(k)}\|_2^2}{\|\mathbf{y}^{(k)}_t - \bar{y}^{(k)}_t\|_2^2},
\end{equation*}
where the subscript \(t\) denotes the subset of pixels belonging to tile \(t\) and \(\bar{y}^{(k)}_t\) their mean. These $R^2$ values quantify how much of the embedding variance the harmonic basis can reconstruct; negative values indicate that the linear model performs worse than simply predicting the mean. We summarise \(R^2_{t,k}\) across tiles using eco-region weights and across embedding channels by averaging over the 14 dimensions. The reported “mean $R^2$” in Table~\ref{tab:similarity_summary} is therefore the eco-region–weighted average over both tiles and embedding channels.

To interpret the fitted coefficients we normalise \(\hat{\boldsymbol{\beta}}^{(k)}\) by their $\ell_1$ norm and record the harmonic descriptor with the largest absolute weight; we also compute Pearson correlations between \(\mathbf{y}^{(k)}\) and each harmonic feature, retaining the largest-magnitude correlation as a complementary cue. Detailed coefficient tables, $R^2$ distributions, and tile-level diagnostics appear in Supplementary Section~S3.

@@

\subsection{H2 -- Spatial coherence and landscape context}
Hypothesis H2 examines whether the embedding representation yields smoother outputs and whether these gains relate to ancillary environmental gradients. Using the densities defined in Section~\ref{sec:metrics}, national edge density drops from 6.69 to 2.76\,km\,km\(^{-2}\) when switching from HARM-14 to EMB-14, a 59\,\% reduction. The median total patch density \(D_{\text{tot}}\) falls from 6.81\,k to 2.16\,k components per 100\,km² (−68\,%), indicating far fewer tiny clusters. The largest smoothing occurs in maritime pine mosaics (Greater Crystalline and Oceanic West), where the harmonic map fragments mixed stands into speckled artefacts. EMB-14 also maintains crisp elevational boundaries in the Alps, Jura, and Vosges, while matching National Forest Inventory proportions (64.4\,\% deciduous, 35.6\,\% evergreen; Figure~\ref{fig:national_map}). Table~\ref{tab:coherence_summary} summarises the national reduction in edge and patch densities, with supplementary Table~S3 providing the full statistics including the median-filter baseline.

Figure~\ref{fig:h2_multiscale} visualises these differences across five representative tiles. Each row pairs a Sentinel-2 chip (left) with the embedding and harmonic predictions (middle and right). The rows cover GRECO letters G (Greater Semi-Continental East), C (Corsica), A (Alps), O (Oceanic Southwest), and the Central Massif. Embeddings preserve large deciduous blocks (amber, \#e3712c) and evergreen belts (azure, \#2693c1) with minimal speckle; the harmonic counterpart produces many 1–2 pixel patches, consistent with its higher \(D_{\text{tot}}\). Additional summaries, including the effect of median filtering, appear in Supplementary Table~S3 and `results/analysis_coherence/coherence_summary.csv`.

@@

Table~\ref{tab:similarity_summary} reports the eco-region–weighted mean of \(R^2_{t,k}\) across tiles and all 14 embedding dimensions, together with the harmonic descriptor that carries the largest normalised ridge weight (the same cue usually matches the top Pearson correlation). Most regions exhibit negative mean $R^2$, confirming that no linear combination of harmonics can reconstruct the embeddings wholesale, even though specific dimensions still correlate with interpretable cues such as CRSWIR offsets or NBR amplitudes. This absence of linear reconstruction, combined with the alignment patterns, supports a hybrid interpretation: AlphaEarth embeddings retain phenological axes while injecting spatial context learned during pre-training. That extra context likely underpins the smoother boundaries and higher accuracy gains observed in mixed Atlantic and montane landscapes, where purely temporal descriptors struggle. Per-dimension coefficients and correlation tables appear in Supplementary Section~S3 for readers who need the fine-grained breakdown.

@@

Calibration quality underpins operational usability. EMB-14 reaches an expected calibration error of 0.033 versus 0.059 for HARM-14, and maintains macro-F1 stability when the decision threshold moves within the \([0.45,0.55]\) range (Supplementary Figure~S6). Most pixels fall either below 0.1 or above 0.9 evergreen probability, and the empirical accuracy curve tracks the diagonal, showing that the forest's stated confidences are largely honest. Where the coloured confidence markers sit above the black empirical markers the model is mildly over-confident (predictions a few points higher than reality); the opposite would signal under-confidence. These deviations never exceed 0.114 (the maximum calibration error), and the average gap is only 0.033 (the expected calibration error), so users can treat the posterior as a direct proxy for evergreen prevalence. A manager who wishes to avoid false evergreen detections can therefore raise the threshold to, say, 0.55, knowing that the selected pixels truly have a \(\gtrsim 55\%\) evergreen prevalence while the overall macro-F1 remains within 0.903–0.906. The remaining calibration gap concentrates around mid-range probabilities (0.4--0.6), which correspond to mixed stands and boundary pixels where either class remains plausible; at the extremes the coloured and black curves overlap, signalling that the forest separates pure deciduous and evergreen pixels cleanly. Full reliability plots and per-threshold confusion tables are provided in Supplementary Figure~S3 and Section~S6. Combined with the reduced speckle visible in Figure~\ref{fig:comparison_products} and the ancillary insights above, these traits simplify both threshold tuning and regional deployment: humid mixed stands can adopt standard thresholds, while dry evergreen regions can either fall back to harmonic cues or tighten evergreen-specific priors without compromising reliability.
